IML Term project paper

### 2022-12-23 paper.rmd

The key objective in this project, classifying NPF event and nonevent days correctly, is an example of *supervised learning*: our data set contains the correct classes, and we can use these in the training of our models. With *unsupervised learning* we can look for insights into the data without using the known classes. There are many different types of unsupervised learning, discussed in [@book, p.497-552], but since our goal is classification, we focus on *cluster analysis*. This means that we look for subgroups of similar observations. In particular, we use K-means clustering to see if we can easily differentiate between event and nonevent days, and between event types Ia, Ib, II and nonevent, without using the given classes except to validate the results. The K-means clustering algorithm starts by assigning each observation to a cluster at random. At every iteration, the algorithm computes a *cluster centroid*, a vector of the variable means for each observation in the *k*th cluster, and then reassigns each observation to the cluster with the nearest centroid. The nearest centroid is the one to which an observation has the smallest Euclidean distance.

Recall that we have daily means and standard deviations in the dataset. For the purposes of K-means clustering, we restrict ourselves to the means, and because the variables are measured on very different scales, we normalize them to have zero mean and unit variance with the *scale* function. The random cluster assignments tend to have an effect on the end result, so we run the *kmeans* function with 1000 random starting points, and look at the best result. With 2 clusters, the confusion matrix given by the *solve_LSAP* function from the [@clue] package that matches the clusters with the known classes, is as follows:

@@

The correlation between the predictors and the value to be predicted is a very natural starting point when it comes to correlation analysis. It is also possible that the predictors themselves are correlated. Many classifiers are affected by high correlation, i.e. *collinearity* between variables. In regression models, for example, it can be difficult to differentiate between the individual effects of collinear variables on the response. Let's visualize these correlations with R package [@corrplot2021]

@@

However, it is not sufficient to simply define the model on the training dataset and assess accuracy and perplexity on said training dataset as this would woefully inflate the accuracy assessment and deflate the perplexity assessment of the model on a general dataset. We instead attempted to assess accuracy and perplexity through cross-validation methods, specifically leave-one-out cross-validation (loo-cv). The idea behind loo-cv according to the course material [@book, p. 200-202] is that for each observation in the training dataset, a model is trained based on all other observations in the dataset. This new model is then used to predict the training dataset value that was left out of defining the model. With this method, we receive based on observations of explanatory variables both predicted values and estimates of probabilities for specific events for each measurement in a way that didn't include the observations in both defining the model training and assessing its quality. Thus, we can use these measures in our accuracy and perplexity calculations to better assess how the model using all the observations in the training dataset behaves on a completely separate test dataset.
```{r preprocess data, echo=FALSE}
# Sample half of the data for training and testing sets
set.seed(42)
idx <- sample(nrow(npfData), nrow(npfData)/2)
npf_train <- npfData[idx, ]
npf_test <- npfData[-idx, ]
```
Next, let's discuss some of the models we considered

@@

After this, we had a remaining set of 18 explanatory variables, which we modelled using logistic regression. After checking the coefficients, there were still variables for which the Wald tests gave relatively high p-values for hypotheses involving non-zero coefficients. In order to deal with this, we set those coefficients equal to zero and assessed the model with the remaining explanatory variables. Through this, we received a simpler model with  explanatory variables.

@@

Lasso regression has obvious links to logistic regression but it includes a penalty term which punishes models with high values of coefficients. According to the course material [@book, p. 241-242] in comparable linear regression minimizing squared error, the minimizable function is the sum of the squared error terms and $\sum_{i=1}^n \gamma |\lambda_i|$ where $\gamma>0$. We similarly as in logistic regression assess coefficients for a model that describes parameter $q$ of a binomial model that minimizes this function.

@@

In logistic regression models, we directly model $p(Y|X)$, i.e. the conditional distribution of the response $Y$ given the predictors $X$. Generative models are covered in [@book, p.141-158]. Generative models offer a less direct approach, where the focus is in modeling $p(X|Y)$ and using these estimates to estimate $p(Y|X)$ for each possible class $k$ with Bayes' theorem:
\begin{equation*}
p(Y=k|X) = \frac{\pi_k p(X|Y=k)}{\sum_i \pi_i p(X|Y=i)}.
\end{equation*}
Here $pi_k$ is an estimate of the prior probability that a random observation comes from the $k$th class, in our case computed as the fraction of the observations in the training dataset that are in the $k$th class.

@@

#### Linear Discriminant Analysis

@@

In our modeling, we used the function lda from the package [@MASS]. We tested the performance of the model on the dataset with observations measured at multiples only observed at 16.8 meters.

#### Quadratic Disriminant Analysis

@@

In our approach, we used the qda function from the [@MASS] package, and as with linear discriminant analysis, tested the performance of the model on the dataset with observations measured at multiples only observed at 16.8 meters.

#### Naive Bayes

A naive Bayes classifier makes no assumptions on the distribution of the observations, and instead assumes that within the *k*th class, the *p* predictors are independent. In other words, the assumption is that there is no association between the predictors, and therefore no joint distribution to consider. With this, the posterior probability can be computed as
\begin{equation*}
p(Y=k|X) = \frac{\pi_k p(X_1|Y=k) \cdots p(X_p|Y=k)}{\sum_i \pi_i p(X_1|Y=i) \cdots p(X_p|Y=i)}.
\end{equation*}
In a naive Bayes model, there are a few possible ways to estimate $p(X|Y)$. For quantitative predictors such as ours, we assume that within each class, the *j*th predictor is drawn from a univariate normal distribution, i.e. $X_j|Y = k \approx N (\mu_{jk}, \sigma^2_jk)$.
The assumption of independence between predictors is often not realistic, but the model can perform well nonetheless. This is especially true for small datasets, since estimating a joint distribution requires a large amount of data. 
We used the naiveBayes function from the package [@e1071] and the original dataset.

@@

#### Random Forest

One often used tree-based classification method is called Random Forest. It is based on the process described above, but only a random subset of the features is considered at each split of the tree. This is repeated multiple times and in the end, the obtained forest is evaluated and the best splits at each point are found. For Random Forest classifier, we used the R library *randomForest*.

@@

#### GBM

Another tree-based model that we used was GBM (gradient boosting machine). It starts by building a set of shallow trees, but with each iteration, it learns more about the data and improves each previously built tree. By finally combining the trees, it can produce very good models for all types of data sets. It's advantage is that it doesn't necessarily require any pre-processing of the data, although the model can be improved by this. For GBM classifier, we used the R library *GBM*.

@@

We also calculated for curiosity's sake the binary accuracies on the npf_test data on some of our models as can be seen above (eventually we came up with the idea very late and thus didn't have time to run it for LDA, QDA and Naive Bayes), even though these results were not part of the process of choosing our model.

>>>>

The key objective in this project, classifying NPF event and nonevent days correctly, is an example of *supervised learning*: our data set contains the correct classes, and we can use these in the training of our models. With *unsupervised learning* we can look for insights into the data without using the known classes. There are many different types of unsupervised learning, discussed in the course material [@book, p.497-552], but since our goal is classification, we focus on *cluster analysis*. This means that we look for subgroups of similar observations. In particular, we use K-means clustering to see if we can easily differentiate between event and nonevent days, and between event types Ia, Ib, II and nonevent, without using the given classes except to validate the results. The K-means clustering algorithm starts by assigning each observation to a cluster at random. At every iteration, the algorithm computes a *cluster centroid*, a vector of the variable means for each observation in the *k*th cluster, and then reassigns each observation to the cluster with the nearest centroid. The nearest centroid is the one to which an observation has the smallest Euclidean distance.

Recall that we have daily means and standard deviations in the dataset. For the purposes of K-means clustering, we restrict ourselves to the means, and because the variables are measured on very different scales, we normalize them to have zero mean and unit variance with the *scale* function. The random cluster assignments tend to have an effect on the end result, so we run the *kmeans* function with 1000 random starting points, and look at the best result. With 2 clusters, the confusion matrix given by the *solve_LSAP* function from the clue[@clue] package that matches the clusters with the known classes, is as follows:

@@

The correlation between the predictors and the value to be predicted is a very natural starting point when it comes to correlation analysis. It is also possible that the predictors themselves are correlated. Many classifiers are affected by high correlation, i.e. *collinearity* between variables. In regression models, for example, it can be difficult to differentiate between the individual effects of collinear variables on the response. Let's visualize these correlations with R package corrplot2021[@corrplot2021]

@@

However, it is not sufficient to simply define the model on the training dataset and assess accuracy and perplexity on said training dataset as this would woefully inflate the accuracy assessment and deflate the perplexity assessment of the model on a general dataset. We instead attempted to assess accuracy and perplexity through cross-validation methods, specifically leave-one-out cross-validation (loo-cv). The idea behind loo-cv according to the course material [@book, p. 200-202], is that for each observation in the training dataset, a model is trained based on all other observations in the dataset. This new model is then used to predict the training dataset value that was left out of defining the model. With this method, we receive based on observations of explanatory variables both predicted values and estimates of probabilities for specific events for each measurement in a way that didn't include the observations in both defining the model training and assessing its quality. Thus, we can use these measures in our accuracy and perplexity calculations to better assess how the model using all the observations in the training dataset behaves on a completely separate test dataset.
```{r preprocess data, echo=FALSE}
# Sample half of the data for training and testing sets
set.seed(42)
idx <- sample(nrow(npfData), nrow(npfData)/2)
npf_train <- npfData[idx, ]
npf_test <- npfData[-idx, ]
```
Next, let's discuss some of the models we considered

@@

After this, we had a remaining set of 18 explanatory variables, which we modeled using logistic regression. After checking the coefficients, there were still variables for which the Wald tests gave relatively high p-values for hypotheses involving non-zero coefficients. In order to deal with this, we set those coefficients equal to zero and assessed the model with the remaining explanatory variables. Through this, we received a simpler model with  explanatory variables.

@@

Lasso regression has obvious links to logistic regression but it includes a penalty term which punishes models with high values of coefficients. According to the course material [@book, p. 241-242], in comparable linear regression minimizing squared error, the minimizable function is the sum of the squared error terms and $\sum_{i=1}^n \gamma |\lambda_i|$ where $\gamma>0$. We similarly as in logistic regression assess coefficients for a model that describes parameter $q$ of a binomial model that minimizes this function.

@@

In logistic regression models, we directly model $p(Y|X)$, i.e. the conditional distribution of the response $Y$ given the predictors $X$. Generative models are covered in the course material [@book, p.141-158]. Generative models offer a less direct approach, where the focus is in modeling $p(X|Y)$ and using these estimates to estimate $p(Y|X)$ for each possible class $k$ with Bayes' theorem:
\begin{equation*}
p(Y=k|X) = \frac{\pi_k p(X|Y=k)}{\sum_i \pi_i p(X|Y=i)}.
\end{equation*}
Here $pi_k$ is an estimate of the prior probability that a random observation comes from the $k$th class, in our case computed as the fraction of the observations in the training dataset that are in the $k$th class.

@@

#### Linear Discriminant Analysis \

@@

In our modeling, we used the function lda from the package MASS[@MASS]. We tested the performance of the model on the dataset with observations measured at multiples only observed at 16.8 meters.

#### Quadratic Disriminant Analysis \

@@

In our approach, we used the qda function from the MASS[@MASS] package, and as with linear discriminant analysis, tested the performance of the model on the dataset with observations measured at multiples only observed at 16.8 meters.

#### Naive Bayes \

A naive Bayes classifier makes no assumptions on the distribution of the observations, and instead assumes that within the *k*th class, the *p* predictors are independent. In other words, the assumption is that there is no association between the predictors, and therefore no joint distribution to consider. With this, the posterior probability can be computed as
\begin{equation*}
p(Y=k|X) = \frac{\pi_k p(X_1|Y=k) \cdots p(X_p|Y=k)}{\sum_i \pi_i p(X_1|Y=i) \cdots p(X_p|Y=i)}.
\end{equation*}
In a naive Bayes model, there are a few possible ways to estimate $p(X|Y)$. For quantitative predictors such as ours, we assume that within each class, the *j*th predictor is drawn from a univariate normal distribution, i.e. $X_j|Y = k \approx N (\mu_{jk}, \sigma^2_jk)$.
The assumption of independence between predictors is often not realistic, but the model can perform well nonetheless. This is especially true for small datasets, since estimating a joint distribution requires a large amount of data. 
We used the naiveBayes function from the package e1071[@e1071] and the original dataset.

@@

#### Random Forest \

One often used tree-based classification method is called Random Forest. It is based on the process described above, but only a random subset of the features is considered at each split of the tree. This is repeated multiple times and in the end, the obtained forest is evaluated and the best splits at each point are found. For Random Forest classifier, we used the R package randomForest[@randomForest].

@@

#### GBM \

Another tree-based model that we used was GBM (gradient boosting machine). It starts by building a set of shallow trees, but with each iteration, it learns more about the data and improves each previously built tree. By finally combining the trees, it can produce very good models for all types of data sets. It's advantage is that it doesn't necessarily require any pre-processing of the data, although the model can be improved by this. For GBM classifier, we used the R library gbm[@gbm].

@@

We also calculated for curiosity's sake the binary accuracies on the npf_test data on some of our models as can be seen above, even though these results were not part of the process of choosing our model.
